---
title: "C03_assignment.Rmd"
author: "Izzy Gottlieb"
date: "2026-01-20"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GitHub Documents

This is an R Markdown format used for publishing markdown documents to GitHub. When you click the **Knit** button all R code chunks are run and a markdown file (.md) suitable for publishing to GitHub is generated.

## Including Code

You can include R code in the document as follows:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
SPECIES2="Eubalaena glacialis"
fetch_obis(SPECIES2)
```

```{r}
obs2=read_obis(SPECIES2)
obs2
```
```{r}
obs2 = obs2 |>
  filter(!is.na(eventDate))
summary(obs2)
```

The next part is making graphs 

```{r plot_month}
ggplot(data = obs2,
       mapping = aes(x = month)) + 
  geom_bar() + 
  labs(title = "Counts per month")
```


```{r}
SPECIES3= "Calanus finmarchicus"
fetch_obis(SPECIES3)
```
```{r}
obs3=read_obis(SPECIES3)
obs3
```

```{r plot_year}
ggplot(data = obs3,
       mapping = aes(x = year)) + 
  geom_bar() + 
  geom_vline(xintercept = c(1982, 2013), linetype = "dashed") + 
  labs(title = "Counts per year")
```


`source("/home/<IzzyTheChromodoris>/<ColbyForecasting_data>/setup.R")`